---
title: "Automated key generation from a sequence alignment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Automated key generation from a sequence alignment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
ok <- requireNamespace("ape", quietly = TRUE) && requireNamespace("phangorn", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = ok)
```

**Automated key generation** builds a dichotomous key from data instead of
from a person's judgement. `vignette("vision-keys")` starts from photos. This
vignette starts from a multiple sequence alignment (MSA) and needs one
command:

```{r}
library(moose)
fasta <- system.file("extdata", "aligned.fasta", package = "moose")   # or the path of your own file
key <- keyFromAlignment(fasta)
```

```{r, eval = FALSE}
exportWizard(key, "key.html")
```

moose ships two example alignments, so that every command here runs as
written:

- `aligned.fasta`: 15 wood mouse cytochrome *b* sequences (the `woodmouse`
  data of the `ape` package), with a tree of them in `aligned.nwk`;
- `mtdna.fasta`: whole human mitochondrial genomes for 14 haplogroups.

## What the command does

1. **Reads the alignment**: a FASTA file, or an object from `ape` or
   `phangorn`.
2. **Builds a tree** with a phylogeny program. The default is neighbour
   joining in `ape`. `tool =` selects another.
3. **Roots the tree** on an outgroup you name, or at its midpoint.
4. **Turns every fork into a couplet.** The two leads of a couplet are the
   alignment sites that tell the two sides of the fork apart, written as the
   site and its state: `106G 35G` against `106A 35A`. Each sequence is a
   result.

The sequences must already be aligned. To align them first, use a program
such as MAFFT or MUSCLE.

## An example

The key built above from `aligned.fasta`, the 15 wood mouse sequences:

```{r}
key
head(key$leads[, c("Statement", "Character", "Next", "Taxon")], 8)
```

Read it like any key. At couplet 1, look at sites 106 and 35 of your
sequence: `G` at both sends you to couplet 2, `A` at both to couplet 3.

The key records how it was made, including the tree itself:

```{r}
meta <- setNames(key$meta$value, key$meta$key)
meta[c("tool", "rooting", "sequences", "sites", "weak_couplets")]
substr(meta[["newick"]], 1, 60)
```

It is an ordinary moose key, so the wizard and the phylogenetics exports work
on it:

```{r, eval = FALSE}
exportWizard(key, "woodmouse.html")
toTreedata(key)                      # for ggtree
```

## Choosing the phylogeny program

```{r, eval = FALSE}
keyFromAlignment(aln, tool = "nj")          # neighbour joining (ape), the default
keyFromAlignment(aln, tool = "bionj")       # BIONJ (ape)
keyFromAlignment(aln, tool = "upgma")       # UPGMA (phangorn)
keyFromAlignment(aln, tool = "parsimony")   # parsimony ratchet (phangorn)
keyFromAlignment(aln, tool = "ml")          # maximum likelihood (phangorn)
keyFromAlignment(aln, tool = "fasttree")    # the FastTree program
keyFromAlignment(aln, tool = "iqtree")      # the IQ-TREE program
```

`"fasttree"` and `"iqtree"` run the command-line program, which must be
installed. moose looks for it on the `PATH`; give its location in `exe` if it
is elsewhere, and extra options in `args`. `model` sets the substitution
model for the tools that take one.

For any other program, build the tree yourself and hand it over. Its tip
labels must be the sequence names. Polytomies are resolved by the sites:

```{r}
newick <- system.file("extdata", "aligned.nwk", package = "moose")    # e.g. RAxML's bestTree file
keyFromAlignment(fasta, tree = newick)
```

## Rooting

A key starts at the root, so where the tree is rooted decides the first
question. Name an outgroup when you have one:

```{r}
data(Laurasiatherian, package = "phangorn")
mammals <- suppressWarnings(keyFromAlignment(Laurasiatherian, outgroup = "Platypus"))
mammals$leads[1:2, c("Character", "Next", "Taxon")]
```

Without an outgroup, an unrooted tree is rooted at its midpoint.

## Numbering sites

By default a site is its column in the alignment. To use the coordinates of
one sequence instead, name it as the `reference`. Columns where the reference
has a gap are numbered as insertions (`2156.1`).

`mtdna.fasta` holds whole mitochondrial genomes for 14 human haplogroups.
One of them, H2a2a1, is the haplogroup of the rCRS, the standard reference
for human mitochondrial DNA, so numbering by it gives the usual coordinates.
L0a1 belongs to the earliest branch of the tree, which makes it the outgroup:

```{r}
mtdna <- system.file("extdata", "mtdna.fasta", package = "moose")
mt <- keyFromAlignment(mtdna, reference = "H2a2a1", outgroup = "L0a1")
mt
mt$leads[1:8, c("Statement", "Character", "Next", "Taxon")]
```

Couplet 4 splits on 8701, 9540 and 10873, the sites that define
macrohaplogroup N, and the other leads are likewise mutations PhyloTree lists
for those branches.

These 14 sequences are not from individuals. Each is rebuilt from PhyloTree
Build 17: the root sequence with the mutations on the path to that haplogroup
applied (see `data-raw/alignments/build.R` in the source).

## When a fork has no clean difference

A site is *diagnostic* for a fork when the states on one side never occur on
the other. A fork can lack such a site, because of homoplasy or because the
tree is wrong there. moose then uses the site that separates the two sides
best, marks the leads `(most)`, and warns:

```{r}
mammals <- keyFromAlignment(Laurasiatherian, outgroup = "Platypus")
grep("(most)", mammals$leads$Character, fixed = TRUE, value = TRUE)
```

If no site separates the two sides of a fork at all, moose stops following
the tree there and splits those sequences by their sites alone, with a
warning. Sequences that no site can tell apart share a result (`"A / B"`), as
do identical sequences. Gaps and ambiguity codes count as missing data and are
never used as a state.

## Identifying a sequence

The key can be followed by machine. `alignmentScores()` reads off the sites
the key tests, and `classify()` walks the key. The new sequences must be
aligned to the same coordinates as the key:

```{r}
classify(key, alignmentScores(key, fasta))[1:4, ]
```

Each couplet is tested on its first listed site. A sequence with a gap or an
ambiguity code at that site gets no result (`NA`).
