---
title: "Mixed-effects conditional inference trees on PhyloTree"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Mixed-effects conditional inference trees on PhyloTree}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
ok <- requireNamespace("partykit", quietly = TRUE) && requireNamespace("lme4", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = ok)
```

A mixed-effects conditional inference tree (MECIT) is a decision tree for
clustered data. It finds splits on candidate variables while a random effect
absorbs differences between clusters. This vignette shows why that matters
for mitochondrial DNA, using expression data simulated on PhyloTree Build 17.

## The problem: same mutation, many clades

Some mtDNA variants are **recurrent**: the same change arose independently on
many branches of the tree. If such a variant changes expression, its effect
shows up in many unrelated lineages. But lineages also differ for reasons that
have nothing to do with that variant. Each large haplogroup (H, U, L3, ...) has
its own background. An ordinary tree cannot tell the two apart. It happily
splits on **clade markers**, variants that simply track which haplogroup a
sample is in, because they explain the haplogroup backgrounds.

`recurrentPositions()` lists the recurrent sites, and `haplotypeMatrix()` says
which haplogroups carry them. Back-mutations are applied the way Haplogrep
defines them, so a site that is gained and later reverted is not carried
below the reversion:

```{r}
library(moose)
head(recurrentPositions(), 8)
haplotypeMatrix(c("H2a2a1", "U5b1", "L0a1b"), sites = c("16362", "150", "16278"))
```

## Simulated data with known truth

`simulateExpression()` samples leaf haplogroups and builds expression in which
a few recurrent sites shift some genes, on top of a random baseline for each
macrohaplogroup. Some genes have only the baseline, and some are pure noise:

```{r}
sim <- simulateExpression(n = 1000, genes = 20, seed = 1)
table(sim$truth$geneType)
sim$effectSites   # sites that can carry real effects
sim$markerSites   # clade markers: no effect, but they track the baselines
```

## Fitting a MECIT

`mecit()` takes the response, the candidate split variables and the clusters.
The default method is RE-EM with conditional inference trees (Fu & Simonoff
2015). It alternates a `partykit::ctree()` on the response, with the
random-effect part removed, and an `lme4::lmer()` fit with one level per
terminal node. Random effects are estimated before the first tree, so that tree
never sees the haplogroup backgrounds:

```{r}
g <- which(sim$truth$geneType == "mutation")[1]
colnames(sim$X)[sim$truth$beta[g, ] != 0]          # the truth

fit <- mecit(sim$expr[, g], sim$X, cluster = sim$samples$macro)
fit
```

The same tree without the random effect also splits on clade markers:

```{r}
plain <- mecit(sim$expr[, g], sim$X, cluster = NULL)
splitVariables(plain)
intersect(splitVariables(plain), sim$markerSites)
```

`method = "mob"` fits the model-based version through
`glmertree::lmertree()` (Fokkema et al. 2018).

## A MECIT is a key

A fitted tree turns into a moose key. Its couplets ask whether the sample
carries each split site, and its results are expression strata. The key
carries machine-readable tests, so `classify()` can place new samples from
their haplotypes, and `exportWizard()` draws it:

```{r}
key <- keyFromMecit(fit)
key$leads[, c("Statement", "Character", "Next", "Taxon")]
head(classify(key, sim$X))
```

```{r, eval = FALSE}
exportWizard(key, "mecit.html")
```

## Random slopes

When a mutation's effect differs between macrohaplogroups, give it a random
slope. The slope variables are also entered as fixed effects, so the random
slopes are deviations around a shared slope rather than a replacement for it:

```{r}
fit2 <- mecit(sim$expr[, g], sim$X, sim$samples$macro,
              slope = colnames(sim$X)[sim$truth$beta[g, ] != 0])
splitVariables(fit2)
```

## References

- Sela RJ, Simonoff JS (2012). RE-EM trees: a data mining approach for
  longitudinal and clustered data. *Machine Learning* 86, 169-207.
- Fu W, Simonoff JS (2015). Unbiased regression trees for longitudinal and
  clustered data. *Computational Statistics & Data Analysis* 88, 53-74.
- Fokkema M, Smits N, Zeileis A, Hothorn T, Kelderman H (2018). Detecting
  treatment-subgroup interactions in clustered data with generalized linear
  mixed-effects model trees. *Behavior Research Methods* 50, 2016-2034.
- van Oven M (2015). PhyloTree Build 17: Growing the human mitochondrial DNA
  tree. *Forensic Science International: Genetics Supplement Series* 5,
  e392-e394.
