---
title: "Building identification keys from images"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building identification keys from images}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Most moose keys are written by people. This vignette shows the other route:
building a key **from a folder of labelled images** with a vision model, the
way the [fordera](https://github.com/leipzig/fordera) project classifies Ford
trucks by model year. moose follows fordera's two methods and generalizes
them to any embedding model.

The design has two layers:

- an **R core** that works on plain numeric matrices of embeddings, so it
  needs no Python and works with any model (`discoverTerms()`,
  `keyFromTerms()`, `keyFromClusters()`, `classify()`, `evaluateKey()`,
  `looKey()`);
- a thin **Python layer**, reached through
  [reticulate](https://rstudio.github.io/reticulate/), that computes CLIP
  image, patch and text embeddings (`visionModel()`, `embedImages()`,
  `embedTexts()`, `scoreImages()`).

Only the embedding step needs Python; everything else is ordinary R.

## Setup

Install the Python side once (CLIP through
[open_clip](https://github.com/mlfoundations/open_clip)):

```sh
pip install -r "$(Rscript -e 'cat(system.file("python/requirements.txt", package="moose"))')"
```

Then load a model. `visionModel()` downloads OpenAI's ViT-B/32 by default, or
takes a local checkpoint:

```{r}
library(moose)
model <- visionModel()                       # or visionModel("ViT-B-32.pt")
```

## An image set

`imageSet()` records one row per image. The `label` argument says which class
each image belongs to; several images can share a label (for example two
photographs of the same species), and a function derives the label from the
file name:

```{r}
images <- imageSet(
  list.files("pics", pattern = "\\.png$", full.names = TRUE),
  label = function(x) sub("_.*$", "", x)      # "1967_alt" -> "1967"
)
```

Embed every image. `embedImages()` returns a global embedding per image and,
by default, the per-patch embeddings a vision transformer produces:

```{r}
emb <- embedImages(model, images$path)
dim(emb$image)     # images x dimensions
dim(emb$patches)   # images x patches x dimensions
```

## Method 1: invented visual terms

`discoverTerms()` clusters every image patch with k-means. Each cluster is a
recurring visual motif, given an invented, pronounceable name because it has
no English one. `termScores()` then scores each image against each term by the
most similar patch, and `keyFromTerms()` grows a key whose every step asks
whether the specimen has one term.

```{r}
terms  <- discoverTerms(emb$patches, k = 40)
scores <- termScores(emb$patches, terms$centroids)
key    <- keyFromTerms(scores, images, terms)
key
```

`keyFromTerms()` offers two tree builders: `"balanced"` (fordera's method,
the default: at each step the term that most evenly splits the remaining
classes) and `"rpart"` (a classification tree, with the split points as
thresholds).

Add example crops so the wizard and the glossary can show what each term
looks like, then export:

```{r}
key <- patchExemplars(model, key, images)
exportWizard(key, "terms.html")
```

The exported page is a step-through wizard with a glossary: each invented
term is defined only by the image patches that match it best.

## Method 2: questions from a vocabulary

The second method keeps the questions in words. You supply a vocabulary of
contrasting visual features; moose clusters the classes, and for each split
picks the vocabulary pair that best separates the two sides, phrased as a
yes/no question.

```{r}
vocab <- data.frame(
  feature  = c("round headlights", "a flat hood", "a chrome bumper"),
  opposite = c("square headlights", "a curved hood", "a painted bumper")
)
txt  <- embedTexts(model, vocabularyPrompts(vocab, "a pickup truck with {x}"))
qkey <- keyFromClusters(emb$image, images, vocab, txt,
                        template = "a pickup truck with {x}")
exportWizard(qkey, "questions.html")
```

## Following a key by machine

A key built this way carries a `features` table describing how to compute
each lead's test from an image, so a machine can walk it. `scoreImages()`
embeds new images and scores them; `classify()` walks the key:

```{r}
s <- scoreImages(key, model, "new-specimen.png")
classify(key, s)
#>            id result        path
#> 1 new-specimen   1967  1a 3b 7a
```

`evaluateKey()` reports accuracy over a labelled set, optionally grouped (for
example model years grouped into design generations). `looKey()` does
leave-one-out evaluation, rebuilding the key without each held-out image:

```{r}
evaluateKey(key, scores, images$label)$accuracy

looKey(images,
  build = function(train) keyFromTerms(scores[train, ], images[train, ], terms),
  score = function(key, i) scores[i, , drop = FALSE])$accuracy
```

With one image per class these figures are optimistic; the leave-one-out
number is the honest one.

## Any model, no Python

Because the core works on matrices, embeddings from any model can be used
directly. If you have an `images x dimensions` matrix `X` and a
`images x patches x dimensions` array `P` from your own encoder, every
function above works without `visionModel()`:

```{r}
terms <- discoverTerms(P, k = 40)
key   <- keyFromTerms(termScores(P, terms$centroids), images, terms)
```

This is how a domain model such as BioCLIP can drive moose for biological
images: embed with that model, then build and evaluate the key in R.
