---
title: "Automated key generation from photos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Automated key generation from photos}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Most keys are written by people. **Automated key generation** builds one from
data instead. This vignette goes from a pile of photos to a dichotomous key;
`vignette("alignment-keys")` does the same from a multiple sequence alignment.

## What you need

- **Photos** of the things you want to tell apart: png, jpg, gif, webp, bmp
  or tiff files. They can be of anything (species, specimens, parts, truck
  model years).
- **A label for each photo** saying which class it shows. A key leads to
  named results, so photos without labels cannot make one. You need at least
  two classes. Several photos per class give a better key than one each.
- **The reticulate package** (1.41 or later). The model that turns photos
  into numbers runs in Python, which reticulate sets up for you (see
  [Setup](#setup)).

## The short version

```{r}
library(dimoose)

pics   <- system.file("extdata", "pics", package = "dimoose")   # example photos; or your own folder
images <- imageSet(pics)                     # 1. photos and their labels
model  <- visionModel()                      # 2. a vision model
emb    <- embedImages(model, images)         # 3. photos -> embeddings
sniglets <- discoverSniglets(emb$patches)    # 4. sniglets: recurring visual features...
scores   <- snigletScores(emb$patches, sniglets$centroids)
key      <- keyFromSniglets(scores, images, sniglets, method = "rpart")   # ...and a key on them
key    <- patchExemplars(model, key, images) # 5. example crops, then a web page
exportWizard(key, "key.html")

key <- renameSniglets(key, c(gaithiark = "a dull, finely textured skin"))   # 6. name what you see
```

The example photos ship with dimoose, so this runs as written: 36 small photos
of six fruits (banana, blueberry, kiwi, lemon, pineapple, strawberry), six of
each, from the [Fruits-360](https://github.com/fruits-360/fruits-360-100x100)
dataset (CC BY-SA 4.0). To use your own photos, put the path of your folder
in place of `pics`.

`key.html` is a self-contained page that walks a person through the key,
showing example crops for each choice. The rest of this vignette explains
each step and the choices in it.

## Setup {#setup}

dimoose uses CLIP through
[open_clip](https://github.com/mlfoundations/open_clip), a Python library,
reached from R with [reticulate](https://rstudio.github.io/reticulate/):

```{r}
install.packages("reticulate")
```

There is nothing to install in Python by hand. The first time you call
`visionModel()`, reticulate sets up a private Python with the packages dimoose
needs (`torch`, `open_clip_torch`, `pillow`, `numpy`) and keeps it for later
sessions. That first call downloads several gigabytes, mostly PyTorch, so
allow it a few minutes.

To use a Python you manage yourself, select it before the first call with
`visionModel(python = "/path/to/python")`, `reticulate::use_virtualenv()` or
the `RETICULATE_PYTHON` environment variable, and install the packages there:

```{r}
reticulate::py_install(c("torch", "open_clip_torch", "pillow", "numpy"))
```

Only step 3 (and the example crops in step 5) needs Python. Everything else is
ordinary R on numeric matrices.

## Step 1: say which photo shows what

`imageSet()` makes a table with one row per photo: an `id`, the file `path`
and the class `label`. There are three common ways photos come labelled, and
it reads all of them.

**One folder per class.** Point `imageSet()` at the top folder. Each photo
takes the name of the folder it is in:

```
pics/
  banana/    0_100.jpg  204_100.jpg  ...
  blueberry/ 0_100.jpg  176_100.jpg  ...
  ...
```

```{r}
pics   <- system.file("extdata", "pics", package = "dimoose")
images <- imageSet(pics)
```

The example photos are laid out this way. The other two layouts are for your
own photos; the folder and file names below are placeholders.

**The class is in the file name.** Give a function that turns a file name
(without its extension) into the label:

```{r}
# robin_01.jpg, robin_02.jpg, wren_01.jpg -> "robin", "robin", "wren"
images <- imageSet("my_photos", label = function(x) sub("_.*$", "", x))
```

**The labels are in a table.** Read the spreadsheet and pass one label per
file:

```{r}
tab    <- read.csv("my_labels.csv")          # columns: file, species
images <- imageSet(file.path("my_photos", tab$file), label = tab$species)
```

Check the result before going on:

```{r}
table(images$label)
#>     banana  blueberry       kiwi      lemon  pineapple strawberry
#>          6          6          6          6          6          6
```

## Step 2: load a vision model

`visionModel()` downloads OpenAI's CLIP ViT-B/32 the first time, or loads a
checkpoint you already have:

```{r}
model <- visionModel()                       # or visionModel("ViT-B-32.pt")
```

For photos of organisms, use BioCLIP instead:
`visionModel(name = "bioclip")`. See [Choosing a model](#models).

The first call also sets up Python (see [Setup](#setup)).

## Step 3: embed the photos

`embedImages()` returns one embedding per photo and one per **tile** of each
photo. Tiles are overlapping windows, 4 x 4 of them by default, each cut out
and embedded on its own as if it were a whole picture:

```{r}
emb <- embedImages(model, images)
dim(emb$image)     # photos x dimensions
#> [1]  36 512
dim(emb$patches)   # photos x tiles x dimensions
#> [1]  36  16 512
```

Tiles are what make the features local. A tile's embedding describes only
what is inside it (a patch of skin, a stem end, an edge), so the same kind of
region can be recognised in photos of different things. In photos with a
plain background, tiles that are mostly background are left out, so the
features are about the subject.

- `tiles = c(2, 4)` adds a coarser grid, for features at two scales.
- `tiles = 0` uses the model's own patch tokens instead. That is quicker, one
  pass per photo, and is what fordera does. But in the last layer of a CLIP
  model each token carries much of the whole picture, so the features found
  tend to say "this is a kiwi", not "this part looks like that part".

This is the slow step: one pass of the model per tile. The later steps take
seconds.

## Step 4: build the key

There are two ways to build the key. They differ in what the questions look
like.

### Sniglets

`discoverSniglets()` clusters all the tiles. Each cluster is a recurring
visual feature: a kind of region that turns up in several photos. It has no
name, so dimoose coins a pronounceable word for it (`gaithiark`) and defines it
by the tiles that match it best. These are **sniglets**. `snigletScores()` scores each photo for each sniglet, and
`keyFromSniglets()` grows a key in which each step asks whether the specimen
has one sniglet:

```{r}
sniglets <- discoverSniglets(emb$patches, k = 40)
scores   <- snigletScores(emb$patches, sniglets$centroids)
key      <- keyFromSniglets(scores, images, sniglets, method = "rpart")
key
#> <dimoose key> Key generated from sniglets
#>   5 couplets, 10 leads, 6 taxa, up to 4 steps deep
```

`keyFromSniglets()` has two tree builders:

- `method = "rpart"` gives each sniglet one threshold (the cut of its scores
  that best separates the classes) and fits a classification tree on
  has/lacks. **Use it when you have several photos per class.**
- `method = "balanced"` (the default, and the method of the
  [fordera](https://github.com/leipzig/fordera) project) picks at each step
  the sniglet that splits the remaining classes most evenly. It is meant for
  **one photo per class**.

On the example photos (six per class), `"rpart"` puts 81% of held-out photos
in the right class and `"balanced"` 58%.

### Questions in words

The second method keeps the questions in plain language. You supply a
vocabulary of contrasting features. dimoose clusters the classes into a tree
and, for each fork, picks the pair that best separates its two sides:

```{r}
vocab <- data.frame(
  feature  = c("a smooth skin", "a round shape", "a yellow colour", "green leaves on top"),
  opposite = c("a rough skin", "a long shape", "a dark colour", "no leaves")
)
template <- "a photo of a fruit with {x}"
txt  <- embedTexts(model, vocabularyPrompts(vocab, template))
qkey <- keyFromClusters(emb$image, images, vocab, txt, template = template)
qkey$leads[1:6, c("Statement", "Character", "Next", "Taxon")]
#>   Statement           Character Next      Taxon
#> 1         1   Has a smooth skin    2
#> 2         1    Has a rough skin    -  pineapple
#> 3         2 Has a yellow colour    3
#> 4         2   Has a dark colour    4
#> 5         3   Has a round shape    - strawberry
#> 6         3    Has a long shape    -     banana
```

The model's answers are not always a person's: here it files the strawberry
under "a yellow colour".

The questions can only be as good as the vocabulary: write pairs a person
could check by eye, and give more pairs than the key has forks.

## Step 5: look at the key and share it

A key is an ordinary dimoose object, so everything else in the package works on
it:

```{r}
key$leads[, c("Statement", "Character", "Next", "Taxon")]
#>    Statement                           Character Next      Taxon
#> 1          1                       Has gaithiark    2
#> 2          1                     Lacks gaithiark    3
#> 3          2      Lacks glienglaund; has laussym    -       kiwi
#> 4          2                     Has glienglaund    -      lemon
#> 5          3                      Lacks glirreur    4
#> 6          3                        Has glirreur    -  pineapple
#> 7          4                      Lacks laissloi    5
#> 8          4                        Has laissloi    - strawberry
#> 9          5 Lacks pruangglaint; has krualdspeus    -     banana
#> 10         5                    Has pruangglaint    -  blueberry
```

Both leads of a couplet describe something. A lead that only said "Lacks
glienglaund" would give the reader nothing to look for, so where it can the
key adds what those photos have instead: a second sniglet that nearly all of
them have and nearly none of the other side does ("Lacks glienglaund; has
laussym"). The machine still tests only the first one.


For a key of sniglets, add example crops so that people can see what each
sniglet means, then export the wizard. Every lead gets examples from the
photos that take it, from as many different classes as do, so they show what
those classes share:

- a "Has" lead shows the sniglet in those photos. For `gaithiark` they are
  the matt skins of kiwis and a lemon;
- a "Lacks" lead shows what its photos have instead: the second sniglet when
  the lead names one, otherwise the regions that come closest to the missing
  one.

The glossary shows each sniglet across all the photos and lists the classes
it is seen in.

```{r}
key <- patchExemplars(model, key, images)
exportWizard(key, "key.html")
```

## Step 6: rename the sniglets

A coined word tells a reader nothing until they have looked at the crops.
Once you have looked, you can say what each sniglet is. `snigletNames()`
lists them:

```{r}
snigletNames(key)
#>         sniglet          name definition  used
#> 1     gaithiark     gaithiark       <NA>  TRUE
#> 2 treuxraikbluk treuxraikbluk       <NA> FALSE
#> 3     zhuprierl     zhuprierl       <NA> FALSE
#> ...
```

Only the sniglets the key uses (`used`) need names. Rename a few directly:

```{r}
key <- renameSniglets(key,
  c(gaithiark = "a dull, finely textured skin"),
  definitions = c(gaithiark = "Matt skin with a fine grain, as on a kiwi or a lemon"))
```

or fill in the table in a spreadsheet and read it back:

```{r}
write.csv(snigletNames(key), "names.csv", row.names = FALSE)
# ... edit the name and definition columns ...
key <- renameSniglets(key, "names.csv")
exportWizard(key, "key.html")
```

The leads now read "Has a dull, finely textured skin", and the glossary shows the new
name, the definition and the coined word it replaced.

How renaming works:

- **The coined word is the sniglet's identifier and never changes.** The name
  is only what people read. Scores, machine tests (`classify()`) and name
  tables saved earlier keep working after a rename.
- **A names table** has the columns `sniglet` (the coined word), `name` and,
  optionally, `definition`. It can be a data frame or a CSV file.
- **A blank name** puts the coined word back.
- **Names must be distinct**, and a name cannot be another sniglet's coined
  word.
- **Write names as noun phrases.** Leads are "Has *name*" and "Lacks *name*".
- **To name sniglets before building keys**, rename the `discoverSniglets()`
  result. Every key built from it then uses the names:
  `sniglets <- renameSniglets(sniglets, "names.csv")`.

## How good is the key?

`evaluateKey()` follows the key by machine for every photo and reports how
many end at the right class. `looKey()` is the honest version: it rebuilds the
key without each photo in turn and then tries that photo.

```{r}
evaluateKey(key, scores, images$label)$accuracy

looKey(images,
  build = function(train) keyFromSniglets(scores[train, ], images[train, ], sniglets, method = "rpart"),
  score = function(key, i) scores[i, , drop = FALSE])$accuracy
#> [1] 0.8055556
```

## Identifying a new photo

The key stores how each of its tests is computed from a photo, so a machine
can follow it. `scoreImages()` embeds new photos and scores them, and
`classify()` walks the key:

```{r}
new <- list.files(system.file("extdata", "new-photos", package = "dimoose"), full.names = TRUE)
classify(key, scoreImages(key, model, new))
#>         id     result        path
#> 1 unknown1     banana 1b 3a 4a 5a
#> 2 unknown2  blueberry 1b 3a 4a 5b
#> 3 unknown3       kiwi       1a 2a
#> 4 unknown4  pineapple       1b 3b
#> 5 unknown5  pineapple       1b 3b
#> 6 unknown6 strawberry    1b 3a 4b
```

Five of the six held-out photos are placed correctly. The fourth is a lemon
that the key files as a pineapple.


## Any model, no Python

The key builders work on matrices, so embeddings from any model can be used
directly. If you have a `photos x dimensions` matrix `X` and a
`photos x regions x dimensions` array `P` from your own encoder, with rows in
the order of `images`, skip steps 2 and 3:

```{r}
sniglets <- discoverSniglets(P, k = 40)
key      <- keyFromSniglets(snigletScores(P, sniglets$centroids), images, sniglets)
```

## Choosing a model: CLIP or BioCLIP {#models}

The model is chosen in step 2, by the `name` you give `visionModel()`:

| Your photos | Model | Call | First download |
|---|---|---|---|
| Anything: objects, products, vehicles, mixed subjects | CLIP ViT-B/32 (OpenAI), the default | `visionModel()` | about 350 MB |
| Organisms: plants, animals, fungi, specimens | [BioCLIP](https://imageomics.github.io/bioclip/) | `visionModel(name = "bioclip")` | about 600 MB |
| Organisms, when accuracy matters more than speed | [BioCLIP 2](https://huggingface.co/imageomics/bioclip-2) | `visionModel(name = "bioclip-2")` | about 1.7 GB |

```{r}
model <- visionModel()                    # regular CLIP
model <- visionModel(name = "bioclip")    # BioCLIP
model <- visionModel(name = "bioclip-2")  # BioCLIP 2
```

- **Only that one line changes.** `embedImages()`, `discoverSniglets()`,
  `keyFromSniglets()`, `patchExemplars()` and `exportWizard()` are called the
  same way with every model.
- **Build and follow a key with the same model.** The key records which
  model made it, and `scoreImages()` warns, or stops, if you give it a
  different one.
- **With BioCLIP, use sniglets.** Its text side was trained on names of
  organisms, not on descriptions of what they look like, so vocabulary
  questions (`keyFromClusters()`) work better with regular CLIP.
- **BioCLIP 2 is a larger model**, so embedding takes several times longer.

BioCLIP is downloaded from the Hugging Face Hub on first use.
`name = "hf-hub:<organisation>/<model>"` loads any other open_clip model
there whose image side is a vision transformer, and
`name = "local-dir:<folder>"` a copy you have already downloaded.
