---
title: "Automated key generation from photos"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Automated key generation from photos}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Most keys are written by people. **Automated key generation** builds one from
data instead. This vignette goes from a pile of photos to a dichotomous key;
`vignette("alignment-keys")` does the same from a multiple sequence alignment.

## What you need

- **Photos** of the things you want to tell apart: png, jpg, gif, webp, bmp
  or tiff files. They can be of anything (species, specimens, parts, truck
  model years).
- **A label for each photo** saying which class it shows. A key leads to
  named results, so photos without labels cannot make one. You need at least
  two classes. Several photos per class give a better key than one each.
- **The reticulate package** (1.41 or later). The model that turns photos
  into numbers runs in Python, which reticulate sets up for you (see
  [Setup](#setup)).

## The short version

```{r}
library(moose)

pics   <- system.file("extdata", "pics", package = "moose")   # example photos; or your own folder
images <- imageSet(pics)                     # 1. photos and their labels
model  <- visionModel()                      # 2. a vision model
emb    <- embedImages(model, images)         # 3. photos -> embeddings
sniglets <- discoverSniglets(emb$patches)    # 4. sniglets: recurring visual features...
scores   <- snigletScores(emb$patches, sniglets$centroids)
key      <- keyFromSniglets(scores, images, sniglets, method = "rpart")   # ...and a key on them
key    <- patchExemplars(model, key, images) # 5. example crops, then a web page
exportWizard(key, "key.html")

key <- renameSniglets(key, c(zhuprierl = "a dimpled yellow peel"))   # 6. name what you see
```

The example photos ship with moose, so this runs as written: 36 small photos
of six fruits (banana, blueberry, kiwi, lemon, pineapple, strawberry), six of
each, from the [Fruits-360](https://github.com/fruits-360/fruits-360-100x100)
dataset (CC BY-SA 4.0). To use your own photos, put the path of your folder
in place of `pics`.

`key.html` is a self-contained page that walks a person through the key,
showing example crops for each choice. The rest of this vignette explains
each step and the choices in it.

## Setup {#setup}

moose uses CLIP through
[open_clip](https://github.com/mlfoundations/open_clip), a Python library,
reached from R with [reticulate](https://rstudio.github.io/reticulate/):

```{r}
install.packages("reticulate")
```

There is nothing to install in Python by hand. The first time you call
`visionModel()`, reticulate sets up a private Python with the packages moose
needs (`torch`, `open_clip_torch`, `pillow`, `numpy`) and keeps it for later
sessions. That first call downloads several gigabytes, mostly PyTorch, so
allow it a few minutes.

To use a Python you manage yourself, select it before the first call with
`visionModel(python = "/path/to/python")`, `reticulate::use_virtualenv()` or
the `RETICULATE_PYTHON` environment variable, and install the packages there:

```{r}
reticulate::py_install(c("torch", "open_clip_torch", "pillow", "numpy"))
```

Only step 3 (and the example crops in step 5) needs Python. Everything else is
ordinary R on numeric matrices.

## Step 1: say which photo shows what

`imageSet()` makes a table with one row per photo: an `id`, the file `path`
and the class `label`. There are three common ways photos come labelled, and
it reads all of them.

**One folder per class.** Point `imageSet()` at the top folder. Each photo
takes the name of the folder it is in:

```
pics/
  banana/    0_100.jpg  204_100.jpg  ...
  blueberry/ 0_100.jpg  176_100.jpg  ...
  ...
```

```{r}
pics   <- system.file("extdata", "pics", package = "moose")
images <- imageSet(pics)
```

The example photos are laid out this way. The other two layouts are for your
own photos; the folder and file names below are placeholders.

**The class is in the file name.** Give a function that turns a file name
(without its extension) into the label:

```{r}
# robin_01.jpg, robin_02.jpg, wren_01.jpg -> "robin", "robin", "wren"
images <- imageSet("my_photos", label = function(x) sub("_.*$", "", x))
```

**The labels are in a table.** Read the spreadsheet and pass one label per
file:

```{r}
tab    <- read.csv("my_labels.csv")          # columns: file, species
images <- imageSet(file.path("my_photos", tab$file), label = tab$species)
```

Check the result before going on:

```{r}
table(images$label)
#>     banana  blueberry       kiwi      lemon  pineapple strawberry
#>          6          6          6          6          6          6
```

## Step 2: load a vision model

`visionModel()` downloads OpenAI's CLIP ViT-B/32 the first time, or loads a
checkpoint you already have:

```{r}
model <- visionModel()                       # or visionModel("ViT-B-32.pt")
```

The first call also sets up Python (see [Setup](#setup)).

## Step 3: embed the photos

`embedImages()` returns one embedding per photo and, by default, one per
patch of each photo (a vision transformer cuts a photo into a grid of
patches):

```{r}
emb <- embedImages(model, images)
dim(emb$image)     # photos x dimensions
dim(emb$patches)   # photos x patches x dimensions
```

This is the slow step; the later ones take seconds.

## Step 4: build the key

There are two ways to build the key. They differ in what the questions look
like.

### Sniglets

`discoverSniglets()` clusters all the patches. Each cluster is a recurring
visual feature. It has no name, so moose coins a pronounceable word for it
(`zhuprierl`) and defines it by the patches that match it best. These are
**sniglets**. `snigletScores()` scores each photo for each sniglet, and
`keyFromSniglets()` grows a key in which each step asks whether the specimen
has one sniglet:

```{r}
sniglets <- discoverSniglets(emb$patches, k = 40)
scores   <- snigletScores(emb$patches, sniglets$centroids)
key      <- keyFromSniglets(scores, images, sniglets, method = "rpart")
key
#> <moose key> Key generated from sniglets
#>   5 couplets, 10 leads, 6 taxa, up to 5 steps deep
```

`keyFromSniglets()` has two tree builders:

- `method = "rpart"` fits a classification tree. **Use it when you have
  several photos per class.**
- `method = "balanced"` (the default, and the method of the
  [fordera](https://github.com/leipzig/fordera) project) picks at each step
  the sniglet that splits the remaining classes most evenly. It is meant for
  **one photo per class**.

On the example photos (six per class), `"rpart"` puts 72% of held-out photos
in the right class and `"balanced"` 44%.

### Questions in words

The second method keeps the questions in plain language. You supply a
vocabulary of contrasting features. moose clusters the classes into a tree
and, for each fork, picks the pair that best separates its two sides:

```{r}
vocab <- data.frame(
  feature  = c("a smooth skin", "a round shape", "a yellow colour", "green leaves on top"),
  opposite = c("a rough skin", "a long shape", "a dark colour", "no leaves")
)
template <- "a photo of a fruit with {x}"
txt  <- embedTexts(model, vocabularyPrompts(vocab, template))
qkey <- keyFromClusters(emb$image, images, vocab, txt, template = template)
qkey$leads[1:6, c("Statement", "Character", "Next", "Taxon")]
#>   Statement           Character Next      Taxon
#> 1         1   Has a smooth skin    2
#> 2         1    Has a rough skin    -  pineapple
#> 3         2 Has a yellow colour    3
#> 4         2   Has a dark colour    4
#> 5         3   Has a round shape    - strawberry
#> 6         3    Has a long shape    -     banana
```

The model's answers are not always a person's: here it files the strawberry
under "a yellow colour".

The questions can only be as good as the vocabulary: write pairs a person
could check by eye, and give more pairs than the key has forks.

## Step 5: look at the key and share it

A key is an ordinary moose object, so everything else in the package works on
it:

```{r}
key$leads[, c("Statement", "Character", "Next", "Taxon")]
#>    Statement           Character Next      Taxon
#> 1          1     Lacks zhuprierl    2
#> 2          1       Has zhuprierl    -      lemon
#> 3          2   Has treuxraikbluk    3
#> 4          2 Lacks treuxraikbluk    -  pineapple
#> 5          3        Has laissloi    -     banana
#> 6          3      Lacks laissloi    4
#> 7          4        Has glirreur    5
#> 8          4      Lacks glirreur    - strawberry
#> 9          5     Lacks floirbaur    -  blueberry
#> 10         5       Has floirbaur    -       kiwi
```

For a key of sniglets, add example crops so that people can see what each
sniglet means, then export the wizard:

```{r}
key <- patchExemplars(model, key, images)
exportWizard(key, "key.html")
```

## Step 6: rename the sniglets

A coined word tells a reader nothing until they have looked at the crops.
Once you have looked, you can say what each sniglet is. `snigletNames()`
lists them:

```{r}
snigletNames(key)
#>         sniglet          name definition  used
#> 1     gaithiark     gaithiark       <NA> FALSE
#> 2 treuxraikbluk treuxraikbluk       <NA>  TRUE
#> 3     zhuprierl     zhuprierl       <NA>  TRUE
#> ...
```

Only the sniglets the key uses (`used`) need names. Rename a few directly:

```{r}
key <- renameSniglets(key,
  c(zhuprierl = "a dimpled yellow peel"),
  definitions = c(zhuprierl = "Bright yellow skin with fine pits, as on a lemon"))
```

or fill in the table in a spreadsheet and read it back:

```{r}
write.csv(snigletNames(key), "names.csv", row.names = FALSE)
# ... edit the name and definition columns ...
key <- renameSniglets(key, "names.csv")
exportWizard(key, "key.html")
```

The leads now read "Has a dimpled yellow peel", and the glossary shows the new
name, the definition and the coined word it replaced.

How renaming works:

- **The coined word is the sniglet's identifier and never changes.** The name
  is only what people read. Scores, machine tests (`classify()`) and name
  tables saved earlier keep working after a rename.
- **A names table** has the columns `sniglet` (the coined word), `name` and,
  optionally, `definition`. It can be a data frame or a CSV file.
- **A blank name** puts the coined word back.
- **Names must be distinct**, and a name cannot be another sniglet's coined
  word.
- **Write names as noun phrases.** Leads are "Has *name*" and "Lacks *name*".
- **To name sniglets before building keys**, rename the `discoverSniglets()`
  result. Every key built from it then uses the names:
  `sniglets <- renameSniglets(sniglets, "names.csv")`.

## How good is the key?

`evaluateKey()` follows the key by machine for every photo and reports how
many end at the right class. `looKey()` is the honest version: it rebuilds the
key without each photo in turn and then tries that photo.

```{r}
evaluateKey(key, scores, images$label)$accuracy

looKey(images,
  build = function(train) keyFromSniglets(scores[train, ], images[train, ], sniglets, method = "rpart"),
  score = function(key, i) scores[i, , drop = FALSE])$accuracy
#> [1] 0.7222222
```

## Identifying a new photo

The key stores how each of its tests is computed from a photo, so a machine
can follow it. `scoreImages()` embeds new photos and scores them, and
`classify()` walks the key:

```{r}
new <- list.files(system.file("extdata", "new-photos", package = "moose"), full.names = TRUE)
classify(key, scoreImages(key, model, new))
#>         id     result           path
#> 1 unknown1     banana       1a 2a 3a
#> 2 unknown2  blueberry 1a 2a 3b 4a 5a
#> 3 unknown3       kiwi 1a 2a 3b 4a 5b
#> 4 unknown4      lemon             1b
#> 5 unknown5  pineapple          1a 2b
#> 6 unknown6 strawberry    1a 2a 3b 4b
```

## Any model, no Python

The key builders work on matrices, so embeddings from any model can be used
directly. If you have a `photos x dimensions` matrix `X` and a
`photos x patches x dimensions` array `P` from your own encoder, with rows in
the order of `images`, skip steps 2 and 3:

```{r}
sniglets <- discoverSniglets(P, k = 40)
key      <- keyFromSniglets(snigletScores(P, sniglets$centroids), images, sniglets)
```

This is how a domain model such as BioCLIP can drive moose for biological
images: embed with that model, then build and evaluate the key in R.
